A Type System for Structured Aliasing in Parallel Programs

### 2012-10-29 doc/types-pldi2013/evaluation.tex
more edits

We evaluate the Legion type system on three criteria:
\begin{itemize}
\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
\item Overhead: Does it reduce checking costs (Section~\ref{subsec:overhead})?
\item Scalability: Can it leverage hierarchical scheduling (Section~\ref{subsec:scalability})?
\end{itemize}
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

These applications show Legion is capable of 
expressing applications with both regular (Fluid, AMR) and irregular (Circuit) pointer data structures.
Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
described are needed in different applications.  Region relationships
are used in all applications. %, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

@@

\subsection{Scalability}
\label{subsec:scalability}
The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
theorem proves it is safe for the runtime to only perform dependence checks between two
tasks that share the same immediate parent task instead of having to consider all
possible interactions with other tasks anywhere in the system.  Since all tasks that
share the same immediate parent are all created on the node where the parent is executing,
all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
Without this property the runtime would be forced to inform all other processors in the system 
of every task that it launches using expensive broadcasts.

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
the overhead of the Legion runtime is always less than 7\% of the total execution time
of the applications.  In some applications communication doesn't scale well, but this is
a result of the algorithm chosen by the programmer, not the Legion runtime.
Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

>>>>

We evaluate the Legion type system on four criteria:
\begin{itemize}
\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
\item Overhead: Does it reduce dynamic checking costs (Section~\ref{subsec:overhead})?
\item Scalability: Can it enable hierarchical scheduling (Section~\ref{subsec:scalability})?
\item Performance: Does the performance increase from relaxed coherence modes 
warrant the increased complexity in the operational semantics (Section~\ref{subsec:relaxedperf})?
\end{itemize}
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

%These applications directly demonstrate the need to for multiple, dynamic partitions
%of program data.
%These applications demonstrate Legion is capable of 
%expressing real applications with both regular (Fluid, AMR) and 
%irregular (Circuit) pointer data structures.
Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
described are needed in different applications.  Region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

@@

\subsection{Scalability}
\label{subsec:scalability}
%The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
%theorem proves it is safe for the runtime to only perform dependence checks between two
%tasks that share the same immediate parent task instead of having to consider all
%possible interactions with other tasks anywhere in the system.  Since all tasks that
%share the same immediate parent are all created on the node where the parent is executing,
%all dependence checks and scheduling decisions can be performed locally with no inter-processor communication.  
%Without this property the runtime would be forced to inform all other processors in the system 
%of every task that it launches using expensive broadcasts.

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
the overhead of the Legion runtime is always less than 7\% of the total execution time
of the applications.  In some applications communication does not scale well, but this is
a result of the algorithm required by the application, not the Legion runtime.
Even in the case of the Circuit application, which exhibits quadratic communication scaling, the Legion
runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

\begin{figure}
\begin{center}
\includegraphics[scale=0.33]{figs/circuit_coherence.pdf}
\end{center}
\vspace{-2mm}
\caption{Performance gains from relaxed coherence modes.\label{fig:circuit_excl}}
\vspace{-6mm}
\end{figure}

\subsection{Performance}
\label{subsec:relaxedperf}

To demonstrate the benefit of relaxed coherence modes, we modified the circuit example from Section~\ref{sec:example}
to use exclusive coherence instead of atomic coherence in the {\em distribute\_charge} task and compared 
the performance of the two versions.
The results are shown in Figure~\ref{fig:circuit_excl}.  Slow-downs ranged from 34\% on 48-GPUs to 67\% on 
96 GPUs and more importantly scaled with node count.  This is a direct consequence of Amdahl's Law.  
Even though the {\em distribute\_charge} tasks are a small fraction of the total work, the serialization 
that results from requiring exclusive access to the 
overlapping ghost regions limits the scalability of the application.  Relaxed coherence modes will be 
imperative for enabling the scalability of applications with aliased data on distributed memory machines.

### 2012-10-29 doc/types-pldi2013/langdef.tex
more edits

\texcomment{
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
permissible.  We do so by proving that any memory traces that can result from the evaluation
of the subexpression must be non-interfering (allowing parallel execution by
Lemma~\ref{lem:noninteffects}).
}
\begin{thm}
\label{thm:hiersched}
\rm
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
non-interfering.
\end{thm}
\texcomment{
\begin{proof}
Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
interleavings, each of which must include all the memory operations of each constituent trace.
By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
\end{proof}
} The Legion task scheduler uses Theorem~\ref{thm:hiersched} in the
following way.  Sibling function calls (those invoked within the same
function body) are checked for non-interference of their (dynamic)
privileges.  If they interfere they are serialized; otherwise they are
considered for execution as parallel subtasks.  Although determining that $e_1$
and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
non-interfering, the converse is not true.  It is possible for $e_1$
and $e_2$ to be judged to be potentially interfering and be serialized
even when the subexpressions are non-interfering.  This loss of
precision can impact performance, and did in a few instances in our
experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
In cases where $e_1$ and $e_2$ are in fact
non-interfering, region partitions can be further refined to make the
disjointness visible in the privileges.  Alternately, the code can be
refactored to lift the subexpressions up to the same level, allowing
the dynamic test based on the region mapping to infer non-interference
in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.

>>>>

\texcomment{
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
permissible.  We do so by proving that any memory traces that can result from the evaluation
of the subexpression must be non-interfering (allowing parallel execution by
Lemma~\ref{lem:noninteffects}).
}
\begin{thm}
\label{thm:hiersched}
\rm
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
non-interfering.
\end{thm}
\texcomment{
\begin{proof}
Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
interleavings, each of which must include all the memory operations of each constituent trace.
By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
\end{proof}
} The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
sibling function calls (those invoked within the same
function body) $e_1$ and $e_2$ are checked for non-interference of their (dynamic)
privileges.  Since $e_1$ and $e_2$ are guaranteed to be created on the node on which their parent
is running, no communication is required to perform the non-interference
test.  If they interfere they are executed in program order or serialized
depending on their coherence specifications; otherwise they are
considered for execution as parallel subtasks.    If $e_1$ and $e_2$ are 
determined to be non-interfering and are scheduled in parallel on different 
remote processors then Theorem~\ref{thm:hiersched}
guarantees that there is no communication required between $e_1$ and $e_2$
to perform non-interference tests between their sub-tasks.  Therefore,
the runtime requires no communication for scheduling.

Although determining that $e_1$
and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
non-interfering, the converse is not true.  It is possible for $e_1$
and $e_2$ to be judged to be potentially interfering and be serialized
even when the subexpressions are non-interfering.  This loss of
precision can impact performance, and did in a few instances in our
experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
In cases where $e_1$ and $e_2$ are in fact
non-interfering, region partitions can be further refined to make the
disjointness visible in the privileges.  Alternately, the code can be
refactored to lift the subexpressions up to the same level, allowing
the dynamic test based on the region mapping to infer non-interference
in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.

### 2012-10-29 doc/types-pldi2013/related.tex
more edits

Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges, 
%and we have reused some DPJ notation, 
there are differences stemming from DPJ's static approach.
Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
partition schemes must be statically decided.  Legion allows 
programmers to create multiple partitions of the same region to give different 
views onto the same data, which is not possible in DPJ.  
%DPJ only supports the 
%equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
%while Legion provides safe execution even in simultaneous environments.  
There is also a difference in emphasis: DPJ requires shared memory, while Legion 
is designed for distributed heterogeneous machines.

@@

Reasoning about disjoint heap data is the strong suit of separation logic \cite{Reynolds02}.  
Concurrent separation logic\cite{Brookes04} has been 
used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we ultimately chose to use a 
privileges system as our primary formalism because separation logic does not easily
support reasoning about the interleaving of operations to non-disjoint regions of memory.

>>>>

Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges, 
%and we have reused some DPJ notation, 
there are differences stemming from DPJ's static approach.
Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
partition schemes must be statically decided.  Legion allows 
programmers to create multiple partitions of the same region to give different 
views onto the same data, which is not possible in DPJ.  DPJ supports both
exclusive and atomic tasks which are similar to Legion's coherence modes, but only
allows specification at the coarser granularity of tasks and not individual regions.
DPJ does not support any modes analogous to simultaneous coherence.
%DPJ only supports the 
%equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
%while Legion provides safe execution even in simultaneous environments.  
There is also a difference in emphasis: DPJ requires shared memory, while Legion 
is designed for distributed heterogeneous machines.

@@

Reasoning about disjoint heap data is the strong suit of separation logic \cite{Reynolds02}.  
Concurrent separation logic\cite{Brookes04} has been 
used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we ultimately chose to use a 
privileges system as our primary formalism because separation logic does not easily
support reasoning about the interleaving of operations to aliased regions of memory.
